---
title: "Building a GraphSpace Object from the Seurat stxBrain Dataset"
author: "Flávio Gabriel Carazza Kessler"
date: "2026-07-06"
bibliography: bibliography.bib
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

# Overview

This vignette demonstrates how to load the `stxBrain` spatial transcriptomics dataset from `SeuratData`, preprocess it with `Seurat` [@Hao2023Seurat], convert the resulting object into a `GraphSpace` object using `RGraphSpace` [@RGraphSpace2026], align the graph coordinates to the tissue image, and save the normalized `GraphSpace` object.

This vignette aims to save the `stxBrain` for the further analysis of cell-cell communication visualization. Please see the vignette: [Plotting cell-cell communication in spatially resolved transcriptomics](https://flaviogckessler.github.io/BioSys_Project/CellChat_ProximityNetwork_PathwaySpace.html). This framework follows the same procedures presented in RGraphSpace: [Spatial Data vignette](https://sysbiolab.github.io/RGraphSpace/articles/spatial-data.html).

# Check required versions

```{r check-required-versions, eval=TRUE, echo=TRUE, message=TRUE, warning=TRUE}
# Check required versions
if (packageVersion("RGraphSpace") < "1.4.1") {
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}

if (packageVersion("PathwaySpace") < "1.4.1") {
  message("Need to update 'PathwaySpace' for this vignette")
  remotes::install_github("sysbiolab/PathwaySpace")
}

if (packageVersion("Seurat") < "5.5.0") {
  message("Need to update 'Seurat' for this vignette")
  remotes::install_github("satijalab/Seurat")
}
```

# Load packages

```{r load-packages, eval=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
library("RGraphSpace")
library("PathwaySpace")
library("Seurat")
library("SeuratObject")
library("SeuratData")
library("patchwork")
library("dplyr")
library("igraph")
```

# Install the Seurat dataset

The `stxBrain` dataset only needs to be installed once. To check which datasets are already installed, run:

```{r check-installed-data, eval=TRUE, echo=TRUE}
SeuratData::InstalledData()

if("stxBrain" %in% SeuratData::InstalledData()$Dataset ){
  message("stxBrain is already installed, no need to install it again.")
}else{
 SeuratData::InstallData("stxBrain") 
}
```

# Load the stxBrain dataset

`LoadData()` may print conversion warnings when loading older SeuratData objects.

```{r load-stxbrain-data, eval=TRUE, echo=TRUE, message=TRUE, warning=TRUE}
seurat_obj <- LoadData("stxBrain", type = "anterior1")
```

# Preprocess the Seurat object

The dataset is normalized using `SCTransform()`, reduced using PCA, and clustered using the standard Seurat workflow. Please see the [spatial transcriptomics vignette](https://satijalab.org/seurat/articles/spatial_vignette) from Seurat.

```{r preprocess-seurat-object, eval=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
seurat_obj <- SCTransform(seurat_obj, assay = "Spatial", verbose = FALSE)

seurat_obj <- RunPCA(seurat_obj, assay = "SCT", verbose = FALSE)

seurat_obj <- FindNeighbors(
  object = seurat_obj,
  reduction = "pca",
  dims = 1:30
)

seurat_obj <- FindClusters(
  object = seurat_obj,
  verbose = FALSE
)
```

# Convert the Seurat object to GraphSpace

The `as.GraphSpace()` function converts the spatial coordinates and associated metadata from the Seurat object into a `GraphSpace` object.

```{r create-graphspace, eval=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
gs <- as.GraphSpace(seurat_obj, space = "spatial",scale = "lowres")
```

# Add the tissue image

If available, the tissue image can be extracted from the Seurat object and assigned to the `GraphSpace` object.

```{r add-tissue-image, eval=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
gs_image(gs) <- SeuratObject::GetImage(seurat_obj, mode = "raster")
```

# Normalize graph coordinates to image space

The graph coordinates are normalized to the image coordinate system. By default, this attempts to align the graph's bottom-up coordinates with the image's top-down matrix layout.

```{r normalize-graphspace, eval=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
gs <- normalizeGraphSpace(gs)
```

# Inspect the GraphSpace object

```{r inspect-graphspace, eval=TRUE, echo=TRUE}
gs
```

# Save the normalized GraphSpace object

The final object is saved as an `.RData` file.

```{r save-graphspace, eval=TRUE, echo=TRUE}
save(gs, file = "stxbrain_Normalized_GraphSpace.RData")
```

# Session information

```{r session-info, eval=TRUE, echo=TRUE}
sessionInfo()
```

# References
